---
title: Estimating epidemiological parameters of Marburg disease in Equatorial Guinea
author: Sebastian Funk
date: 31 March, 2023
output: github_document
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  message = FALSE,
  warning = FALSE,
  fig.width = 10,
  fig.height = 7
)
```

```{r init, include = FALSE}
library("googlesheets4")
library("dplyr")
library("janitor")
library("cmdstanr")
library("posterior")
library("tidyr")
library("ggplot2")
library("lubridate")
library("here")
library("knitr")
devtools::load_all(here::here())

linelist <- googlesheets4::read_sheet(
  paste0("https://docs.google.com/spreadsheets/d/1iviEt9eZCPDNjMSEqIfLXl0W2LlHXrt93Z6FI_1seAM/edit#gid=0"), ## nolint
  n_max = 31) |>
  janitor::clean_names()

wiw <- googlesheets4::read_sheet(
  paste0("https://docs.google.com/spreadsheets/d/1iviEt9eZCPDNjMSEqIfLXl0W2LlHXrt93Z6FI_1seAM/edit#gid=0"), ## nolint
  sheet = "Who_Infected_Who",
  n_max = 14) |>
  janitor::clean_names()

null_death <- which(vapply(linelist$date_death, is.null, FALSE))
linelist[null_death, ]$date_death <-
  lapply(null_death, function(x) "NA")

linelist <- linelist |>
  dplyr::rowwise() |>
  dplyr::mutate(
    date_onset =
      lubridate::ymd(as.character(unlist(date_onset))),
    date_death =
      lubridate::ymd(as.character(unlist(date_death))),
  ) |>
  dplyr::ungroup()

wiw <- wiw |>
  mutate(
    from = as.integer(sub("^ID", "", from)),
    to = as.integer(sub("^ID", "", to))
  )
```

# Summary

Initial exploration of serial intervals and distribution from onset to death for Marburg disease in Equatorial Guinea, using data from a [published report](https://www.guineasalud.org/archivos/Ordenes/Comunicado3.pdf).

```{r run_model, include = FALSE}
model <- cmdstanr::cmdstan_model(here::here("inst/stan/model.stan"))

data <- marburg::prepare_data(linelist, wiw, init = 21)
t0 <- min(linelist$date_onset, na.rm = TRUE) - 21

init <- marburg::prepare_init(data)

fit <- model$sample(
  data = data,
  init = init,
  chains = 2,
  parallel_chains = 2,
  adapt_delta = 0.99
)

draws <- fit$draws() |>
  posterior::as_draws_df() |>
  tidyr::pivot_longer(c(-.chain, -.iteration, -.draw))
  
summary <- fit$summary()
```

# Model description

We estimate times of onset and death for all `r nrow(linelist)` reported cases. We further estimate the parameters of the onset-to-death distribution, and the serial interval from `r nrow(wiw)` established transmission pairs.

We modelled serial intervals as strictly non-negative and drawn from a gamma distribution with parameters estimated from the data. We modelled the onset-to-death distribution as drawn from a lognormal distribution with parameters estimated from the data. In obtaining the parameters we accounted for daily censoring and truncation at the end of the observation period.


# Parameters of estimated distributions

**These results have been derived without a full understanding of the data collection process and therefore need to be treated with caution.**

```{r summaries, echo = FALSE}
estimate <- list()
estimate[["Serial interval (mean)"]] <- summary |>
  dplyr::filter(variable == "si_mean")
estimate[["Serial interval (standard deviation)"]] <- summary |>
  dplyr::filter(variable == "si_sd")
estimate[["Delay from onset to death (median)"]] <- summary |>
  dplyr::filter(variable == "od_median")
estimate[["Delay from onset to death (standard deviation)"]] <- summary |>
  dplyr::filter(variable == "od_sd")
estimate <- estimate |>
  dplyr::bind_rows(.id = "Parameter") |>
  dplyr::mutate(
    Estimate = paste0(round(mean), " days (95% CI: ", round(q5), " -- ", round(q95), ")")
  )
estimate |>
  dplyr::select(Parameter, Estimate) |>
  knitr::kable()
```


# Estimated curve of symptom onsets

```{r epicurve, echo = FALSE}
epicurve <- draws |>
  dplyr::filter(grepl("^time_onset", name)) |>
  tidyr::separate(name, c("name", "index"), sep = "\\[") |>
  dplyr::mutate(
    index = as.integer(sub("\\]", "", index)),
    day = as.integer(value)
  ) |>
  dplyr::group_by(.draw, day) |>
  dplyr::summarise(cases = n(), .groups = "drop") |>
  dplyr::group_by(.draw) |>
  tidyr::complete(day = seq_len(max(day))) |>
  dplyr::ungroup() |>
  tidyr::replace_na(list(cases = 0)) |>
  dplyr::mutate(
    date = t0 + day,
    week = lubridate::floor_date(date, unit = "week", week_start = 1)
  ) |>
  dplyr::group_by(.draw, week) |>
  dplyr::summarise(
    cases = sum(cases),
    n = n(),
    .groups = "drop"
  ) |>
  dplyr::filter(n == 7)

p <- ggplot2::ggplot(epicurve,
  ggplot2::aes(x = week, y = cases, group = .draw)) +
  ggplot2::geom_line(alpha = 0.2) +
  ggplot2::theme_bw() +
  ggplot2::xlab("Week (final day)") +
  ggplot2::ylab("Estimated number of cases")
print(p)
```
